### 2013-05-14 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
refactoring spn chapter sections
adding mcsa breakdown data spreadsheet

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10uspj}
\end{table}
It is readily apparent from the tabulated data that point Jacobi
preconditioning is insufficient. For problems of order $SP_5$ and
$SP_7$, the method will not converge at all and for the upscatter case
with $SP_3$ the spectral radius is still quite large for large $P_N$
orders and therefore convergence is expected to be slow. These
eigenvalues signal a need for a better preconditioning strategy to
both ensure and improve covergence for Monte Carlo methods.

@@

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we instead
choose \textit{block Jacobi} preconditioning as a left preconditioner
for the system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}. For the $SP_N$ equations,
Eq~(\ref{eq:A_block_matrix}) gives a block size of $N_g\times(N+1)/2$.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
The inversion of this preconditioner is trivial as shown on the right
side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
individually and combined to form the inverse. In addition, in the
limit of a block size of one, the block Jacobi method reduces to the
point Jacobi method. For high performance implementations this has
several attractive properties. First, the blocks in the matrix come
from the energy/angle discretization of the transport equation as
given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
bound to a mesh element in the system (note there are 8 blocks on the
diagonal in each of the sparsity patterns with a mesh of $2 \times 2
\times 2$) and therefore we expect the matrix elements forming the
block to be entirely local. Second, these blocks are typically dense
and nearly lower triangular for many transport problems meaning that
established dense matrix methods can be used for fast inversion.

@@

From the tabulated block Jacobi data it is clear that this is a viable
preconditioning choice for all $SP_N$ problems in term of Monte Carlo
solution methods. All cases were observed to have a spectral radius
below unity and often significantly smaller than that, greatly
improving convergence rates over the point Jacobi preconditioned
problem. Based on these results, the block Jacobi method should be the
first preconditioning strategy applied to more complicated problems
representative of physical systems.

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Each fuel pin is resolved by a $2 \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly.}}
  \label{fig:problem3_axial_mat}
\end{figure}
Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, core
plenum, and moderator with boron. Table~\ref{tab:problem3_parameters}
gives the primary design parameters for the fuel assembly
calculations. Group cross sections and other discrete nuclear data are
generated as needed by a cross-section processing module dependent on
the meshing parameters used to discretize the geometry.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate converge and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

\subsection{Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Even if the problem may be
ill-conditioned, we do expect convergence of MCSA. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same problem. Figure~\ref{fig:block_jacobi_res_richardson}
gives the residual infinity norm as a function of iteration for the
Richardson linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.}  \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration. However,
convergence is achieved meaning that the criteria needed to satisfy
MCSA convergence has also been met. The effective slope of the
convergence curve in Figure~\ref{fig:block_jacobi_res_richardson} will
give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). For the data in
Figure~\ref{fig:block_jacobi_res_richardson}, $\rho(\mathbf{H})
\approx 0.985$, nearing the limits of MCSA applicability and well
beyond the spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used. Based on this, the correction being
generated by the Neumann-Ulam solve must be incorrect.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi calculations
used a one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use a simpler homogoneous 2-dimensional
one-speed neutron diffusion system to isolate this behavior. In this
system, we can vary the cross sections in order to achieve varying
spectral radii.

@@

\subsection{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}

\subsubsection{ILUT Preconditioning}
\label{subsubsec:spn_ilut_preconditioning}

\subsubsection{Sparse Approximate Inverse Preconditioning}
\label{subsubsec:spn_spainv_preconditioning}

\subsubsection{SPAINV Improved ILUT Preconditioning}
\label{subsubsec:spn_spainv_preconditioning}

\subsubsection{MCSA Relaxation Parameters}
\label{subsubsec:spn_mcsa_relaxation}

\subsubsection{Monte Carlo Estimator Comparison}
\label{subsubsec:spn_estimator_comparison}

\subsection{MCSA Comparison to Conventional Methods}
\label{subsec:spn_comparison}

>>>>

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10uspj}
\end{table}
It is readily apparent from the tabulated data that point Jacobi
preconditioning is insufficient. For problems of order $SP_5$ and
$SP_7$, the method will not converge at all and for the upscatter case
with $SP_3$ the spectral radius is still quite large for large $P_N$
orders and therefore convergence is expected to be slow. These
eigenvalues signal a need for a better preconditioning strategy to
both ensure and improve convergence for Monte Carlo methods.

@@

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we instead
choose \textit{block Jacobi} preconditioning as a left preconditioner
for the system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
trivial with each diagonal block inverted separately. For the $SP_N$
equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
$N_g\times(N+1)/2$.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
The inversion of this preconditioner is trivial as shown on the right
side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
individually and combined to form the inverse. In addition, in the
limit of a block size of one, the block Jacobi method reduces to the
point Jacobi method. For high performance implementations this has
several attractive properties. First, the blocks in the matrix come
from the energy/angle discretization of the transport equation as
given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
bound to a mesh element in the system (note there are 8 blocks on the
diagonal in each of the sparsity patterns with a mesh of $2 \times 2
\times 2$) and therefore we expect the matrix elements forming the
block to be entirely local. Second, these blocks are typically dense
and nearly lower triangular for many transport problems meaning that
established dense matrix methods can be used for fast inversion.

@@

From the tabulated block Jacobi data it is clear that this is a viable
preconditioning choice for all $SP_N$ problems in term of Monte Carlo
solution methods. All cases were observed to have a spectral radius
below unity and often significantly smaller than that, greatly
improving convergence rates over the point Jacobi preconditioned
problem. Based on these results, the block Jacobi method should be the
first preconditioning strategy applied to more complicated problems
representative of physical systems.
\clearpage

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. On the top and bottom of the
assembly, vacuum conditions are used as well as on the top and right
boundaries in Figure~\ref{fig:problem3_radial_mat}. Reflecting
conditions are used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenzied moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly.}}
  \label{fig:problem3_axial_mat}
\end{figure}
Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, core
plenum, and moderator with boron. Group cross sections and other
discrete nuclear data are generated as needed by a cross-section
processing module dependent on the meshing parameters used to
discretize the geometry and single-dimension pin-cell calculations for
intial flux spectrum generation. Table~\ref{tab:problem3_parameters}
gives the primary design parameters for the fuel assembly
calculations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation\footnote{The operator will change if, for example, physics
  feedback through temperature or potentially burnup is
  considered. These additional physics will modify the cross sections
  used to assemble $\mathbf{M}$. The calculations presented here will
  consist of a single eigenvalue calculation with a static
  $\mathbf{M}$.} while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Even if the problem may be
ill-conditioned, we do expect convergence of MCSA. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same problem. Figure~\ref{fig:block_jacobi_res_richardson}
gives the residual infinity norm as a function of iteration for the
Richardson linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.}  \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the criteria needed to satisfy
MCSA convergence has also been met. The effective slope of the
convergence curve in Figure~\ref{fig:block_jacobi_res_richardson} will
give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). For the data in
Figure~\ref{fig:block_jacobi_res_richardson}, $\rho(\mathbf{H})
\approx 0.985$, nearing the limits of MCSA applicability and well
beyond the spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used for these preliminary
calculations. Based on this, the correction being generated by the
Neumann-Ulam solve is the component of MCSA causing a divergent
solution.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi
preconditioned\footnote{For 1-group $SP_1$ problems, the block size is
  1, giving a point Jacobi preconditioner.}  calculations used a
one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use a simpler homogoneous 2-dimensional
one-speed neutron diffusion system to isolate this behavior. In this
system, we can vary the cross sections while maintaining a fixed grid
in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

@@

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}

\subsection{SPAINV Improved ILUT Preconditioning}
\label{subsec:spn_spainv_preconditioning}

\subsection{MCSA Relaxation Parameters}
\label{subsec:spn_mcsa_relaxation}

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Conventional Methods}
\label{sec:spn_comparison}

%%---------------------------------------------------------------------------%%
\section{Multiple Fuel Assembly Challenge Problem}
\label{sec:spn_challenge_problem}
